Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-10 aistats.tex
m

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{figs/}}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator

@@

@@

Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.

\fdk{Online learning;
\citet{azoury2001relative,freund1996predicting,dasgupta2007online}
}

More recently, advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
relative to a reference function $A$ is defined  as 
\aligns{
	\mu \cB_{A}(x, y)
	\leq 
	\cB_f(x,y)
	\leq 
	L \cB_A(x,y).
}
Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
 Exponential families log-likelihood perfectly fits into this theory.
\aligns{
	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates 
for all exponential families.

{\bf Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.

Relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
Recent works have made attempts to generalize the assumptions 
that leads to the convergence of mirror descent 
with strongly-convex reference functions, 
such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
$\expect{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
or only requiring this bounded variance at the minimum $\theta_*$.
However, they do not yet address the problem satisfactorily.
We present those generalization here 
and they are summarized in the context of maximum likelihood estimation 
in \cref{tbl:assumptions}.

@@

@@

\subsection{Bounded variance}
An adaptation of the bounded variance,
investigated by \citet{hanzely2018fastest}, 
defines the following bounded quantity.
Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
\alignn{\label{eq:assumption-hanzely}
	\expect[\tilde g]{
	\lin{
		\nabla f(\theta) - \tilde g(\theta), 
		\theta^+\!\!\paren{\eta \tilde g(\theta)} - \theta^+\!\!\paren{\eta \nabla f(\theta)}
	}
	}
	\leq 
	\eta C,
}
for some step-size $\eta$ and constant $C$. Let us unpack this quantity;

{\bf For a Gaussian} with known variance, 
with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
\aligns{
	\frac{1}{\mu} \expect{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*},
}
and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.

{\bf In divergence,} this condition states that the (symmetrized) divergence between 
a deterministic and stochastic update is bounded in expectation,
\alignn{\label{eq:hanzely-in-divergence}
	\begin{aligned}
	&
	\mathbb{E}_{\tilde g}\bigg[
	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
	\\
	&\hphantom{\mathbb{E}_{\tilde g}[}
	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
	\bigg]
	\leq 
	\eta C.
	\end{aligned}
}

{\bf For a maximum likelihood problem, }
where $\tilde g(\theta) = \m - x$ 
where $x \sim p(x\cond\theta_*)$,
and $\nabla f(\theta) = \m - \m_*$,
\aligns{
	&
	\mathbb{E}_{x}\bigg[
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
	\\
	&\hphantom{\mathbb{E}_{\tilde g_t}[}
	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\bigg]
	\leq 
	\eta C.
}
\fdk{Could work with decreasing step-sizes}

@@

\textbf{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}

\bibliographystyle{apalike}
\bibliography{references.bib}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{figs/}}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand*{\expecti}[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator

@@

\newpage~\null\\
\FloatBarrier

@@

Mirror Descent (MD)\footnote{%
Also known as is going by many names: 
Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.
} 
\citep{nemirovski1983problem} 
was first introduced to optimize non-smooth functions \fdk{(Is typically encountered in non-smooth, online optimization?)},
typically under bounded gradient assumption on the objective $f$ 
or strong convexity assumption on the potential $\logpart$

\fdk{
{\bf Online learning paragraph?}
\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
Bubeck's paragraph
}

More recently, advances in optimization that go beyond the Euclidean case, 
to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
relative to a reference function $A$ is defined  as 
\aligns{
	\mu \cB_{A}(x, y)
	\leq 
	\cB_f(x,y)
	\leq 
	L \cB_A(x,y).
}
Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
Exponential families log-likelihood perfectly fits into this theory.
\aligns{
	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.
However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates 
for all exponential families.

@@

{Traditional results}
in stochastic optimization,
such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
rely on the assumption that the reference function $A$ 
is itself strongly convex relative to a norm $\norm{\cdot}$,
(meaning that $f$ is strongly convex). 
This apply for the MLE problem of a Gaussian distribution 
with fixed variance, 
but not in a general case. 
when the covariance is unknown,
or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
are not strongly convex.

The introduction of relative smoothness has renewed interest for analyses of mirror descent
without relying on strong-convexity in an underlying norm.
Recent works have made attempts to generalize the assumptions 
that are used to control the convergence of stochastic gradient descent 
to this setting. 
Those include 
bounded gradients $(\expecti{\norm{\tilde g(\theta)}^2} \leq C$, 
bounded variance $(\expecti{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq C$
uniformly or only at the minimum $\theta_*$,
or bounded disparities between the minimum of $f$ at $\theta_*$ and the minima of individual functions $f_i$, 
if $f(\theta) = \expect[i]{f_i(\theta)}$ $(\expect{f(\theta_*) - \min_\theta f_i(\theta)} < C)$.
(\fdk{For a review of those assumptions, see that paper by Richtarik's students})

However, they do not yet address the problem satisfactorily.
We present those generalization here 
and they are summarized in the context of maximum likelihood estimation 
in \cref{tbl:assumptions}.

@@

\subsection{Bounded variance}
An adaptation of the bounded variance,
investigated by \citet{hanzely2018fastest}, 
defines the following bounded quantity.
Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
\alignn{\label{eq:assumption-hanzely}
	\expect[\tilde g\!]{
	\lin{
		\nabla\! f(\theta) 
		- \tilde g(\theta), 
		\theta^+\!\!\paren{\eta \tilde g(\theta)} 
		- \theta^+\!\!\paren{\eta \nabla\! f(\theta)}
	}
	}
	\leq 
	\eta C,
}
for some step-size $\eta$. Let us unpack this quantity;

{\bf For a Gaussian} with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$), 
this definition recovers the variance of the stochastic gradient
$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
the LHS of \cref{eq:assumption-hanzely} is bounded by 
$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
to hold.

{\bf In divergence,} this condition states that the (symmetrized) divergence between 
a deterministic and stochastic update is bounded in expectation,
\alignn{\label{eq:hanzely-in-divergence}
	\begin{aligned}
	&
	\mathbb{E}_{\tilde g}\bigg[
	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
	\\[-1em]
	&\hphantom{\mathbb{E}_{\tilde g}[}
	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
	\bigg]
	\leq 
	\eta C.
	\end{aligned}
}
{\bf For a maximum likelihood problem, }
where $\tilde g(\theta) = \m - x$ 
where $x \sim p(x\cond\theta_*)$,
and $\nabla f(\theta) = \m - \m_*$,
\aligns{
	&
	\mathbb{E}_{x}\bigg[
	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
	\\
	&\hphantom{\mathbb{E}_{\tilde g_t}[}
	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
	\bigg]
	\leq 
	\eta C.
}
\fdk{Could work with decreasing step-sizes}

@@

\textbf{Bounded optimality gap,}
\fdk{
Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
\\
Do they do something with that? It seems to be fixed step-size only.
}
\aligns{
	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
}

\subsection{Issues with current analyses}

\fdk{They don't hold for problems we care about}

\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}

\newpage
\bibliographystyle{apalike}
\bibliography{references.bib}
